---
title: "Get started"
author: Konrad Rudolph
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
    md_document:
        variant: gfm
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{Get started}
    %\VignetteEncoding{UTF-8}
---
```{r include = FALSE}
devtools::load_all()
detach('package:box') # No need to keep the package attached.

# The module does not export any symbols. We use `[...]` solely to prevent the
# declaration to create a module alias in the calling environment, so this
# implementation detail isn’t shown in subsequent `ls()` calls.
box::use(./source_file[...])
```

## Using modules

For the purpose of this tutorial, we are going to use the example module
`bio/seq`. The module implements some very basic mechanisms for dealing with DNA
sequences (= character strings consisting of the letters `A`, `C`, `G` and `T`).

First, we load the module:

```{r}
box::use(./bio/seq)
```

The function `box::use` accepts a list of *unquoted*, *qualified* module names.
Each of these module names will load a single module and make it available to
the caller in some form. In the code above, we’ve loaded a single module,
`seq/seq`. `bio` serves as a *parent module* that may group several submodules.
Since the module name inside `box::use` starts with `./`, the module location is
resolved *locally*, i.e. relative to the path of the currently running code.

In the above, `seq` is the module’s *proper name*. `bio/seq` is its *fully
qualified name*. And `./bio/seq` is its *`use` declaration*.

To see the effect of this `use` declaration, let’s inspect our workspace:

```{r}
ls()
seq
```

We have used the module’s fully qualified name to load it. But, as shown by
`ls`, loading the module this way only introduces a single new name into the
current scope, the module itself, identified by its proper (non-qualified) name.

To see which names a module exports, we use `ls` again, this time on the module
itself:

```{r}
ls(seq)
```

It appears that `seq` exports `r length(ls(seq))` different names. To access
exported names, we use the `$` operator:
<code>`r paste0('seq$', lsf.str(seq)[[1L]])`</code> allows us to use the first
function in the list of exported names. We can also display the interactive help
for individual names using the `box::help` function, e.g.:

```{r eval = FALSE}
box::help(seq$revcomp)
```

Now let’s actually *use* the module. The `seq` function inside the `dna/seq`
module constructs a set of (optionally named) biological sequences:

```{r}
s = seq$seq(
    gene1 = 'GATTACAGATCAGCTCAGCACCTAGCACTATCAGCAAC',
    gene2 = 'CATAGCAACTGACATCACAGCG'
)

seq$is_valid(s)

s
```

Note how we automatically get pretty-printed ([FASTA][]) output because the
`print` method (which gets called implicitly here) is specialised for the
`'bio/seq'` S3 class in the `bio/seq` module (prefixing S3 classes inside
modules with the full module name is a convention to avoid name clashes of S3
classes):

```{r}
getS3method('print', 'bio/seq')
```

The source code for `print.bio/seq` contains an interesting `use` declaration.
It show-cases an alternative way of invoking `box::use`, which we’ll explore
now.

## Attaching modules

Let’s have a look at alternative ways of using modules.

To start, let’s unload the `bio/seq` module …

```{r}
box::unload(seq)
```

… and load it again, via a different route:

```{r}
options(box.path = getwd())
box::use(bio/seq[revcomp, is_valid])
```

After unloading the already loaded module, `options(box.path = …)` sets the
module search path: this is where `box::use` searches for modules. If more than
one path is given, `box::use` searches them all until a module of matching name
is found. This works analogously to how `.libPaths` operates on R packages.

The `box::use` directive can now use `bio/seq` instead of `./bio/seq` as the
module name: rather than a relative name we specify a *global* name. In this
example we set the search path to the current working directory but in normal
usage it would be a global library location, e.g. (following the [XDG base
directory specification][XDG]) `~/.local/R/modules` on Linux.

[FASTA]: https://en.wikipedia.org/wiki/FASTA_format
[XDG]: https://specifications.freedesktop.org/basedir-spec/basedir-spec-latest.html

**Note** that non-local module names *must* be fully qualified, nested modules:
`box::use(foo/bar)` works, `box::use(bar)` does not (instead, it is assumed that
`bar` refers to a *package*)!

In the declaration above we use `[revcomp]` to specify that the `revcomp` name
from the `bio/seq` module should be attached in the calling environment. The
`[…]` part is an *attach specification*: a comma-separated list of names inside
the parentheses specifies which names to attach. The special symbol `...`
specifies that *all exported names* should be attached. This has an effect
similar to conventional package loading via `library` (or `attach`ing an
environment): all the attached names are now available for direct use without
necessitating the `seq$` qualifier:

```{r}
is_valid(s)
revcomp(s)
```

However, unlike the `attach` function, module attachment happens in the
*current, local scope* only.

Since the above code was executed in the global environment, there’s no
distinction between local and global scope:

```{r}
search()
```

Note the second item, which reads “<code>`r search()[2L]`</code>”. But let’s now
undo that, to attach (and use) the module locally instead:

```{r}
detach()

seq_table = function (s) {
    box::use(./bio/seq[...])
    table(s)
}

seq_table(s)
```

Unlike above, we are now attaching *all* exported names instead of specifying
individual names. The subsequent line of code uses the `seq$table` function
rather than `base::table` (which would have a different output). And note that
the `seq` module’s `table` function is *not* attached outside the local scope:

```{r}
search()
table(s)
```

This is very powerful, as it isolates separate scopes more effectively than the
`attach` function. What is more, modules which are used and attached inside
another module *remain* inside that module and are not visible outside the
module by default.

Nevertheless, the normal, recommended usage of a module is without an attach
specification, as this makes it clearer which names are being referring to.

## Writing modules

The module `bio/seq`, which we have used in the previous section, is implemented
in the file [`bio/seq.r`](bio/seq.r). The file `seq.r` is, by and large, a
normal R source file, which happens to live in a directory named `bio`.

In fact, there are only three things worth mentioning:

1. Documentation. Functions in the module file can be documented using
   ‘[roxygen2](https://cran.r-project.org/package=roxygen2)’
   syntax. It works the same as for packages. The ‘box’ package parses the
   documentation and makes it available via `box::help`. *Displaying module help
   requires that ‘roxygen2’ is installed.*

2. Export declarations. Similar to packages, modules explicitly need to declare
   which names they export; they do this using the annotation comment `#'
   @export` in front of the name. Again, this works similarly to ‘roxygen2’ (but
   does *not* require having that package installed).

3. The module exports [S3 functions](http://adv-r.had.co.nz/S3.html). The
   ‘box’ package takes care to register such functions automatically but
   this only works for *user generics* that are defined inside the same module.
   When overriding “known generics” (such as `print`), we need to register
   these manually via `register_S3_method` (this is necessary since these
   functions are inherently ambiguous and there is no automatic way of finding
   them).

## Nesting modules

Modules can also form nested hierarchies. In fact, here is the implementation of
`bio` (in [`bio/__init__.r`](bio/__init__.r): since `bio` is a directory rather
than a file, the module implementation resides in the nested file `__init__.r`):

```{r file = 'bio/__init__.r'}
```

The submodule is specified as `./seq` rather than `seq`: the explicitly provided
relative path prevents lookup in the import search path (that we set via
`options(box.path = …)`); instead, only the current directory (that is, the
directory containing the `bio` module) is considered.

We can now use the `bio` module:

```{r}
options(box.path = NULL) # Reset search path
box::use(./bio)
ls(bio)
ls(bio$seq)
bio$seq$revcomp('CAT')
```

We could also have implemented `bio` as follows:

```{r eval = FALSE}
#' @export
box::use(./seq[...])
```

This would have made all of `seq`’s definitions immediately available in
`bio`, without having to always write `seq$…`. This is sometimes useful, but
should be employed with care: being explicit about namespaces generally
increases code robustness and readability.

## Code execution on loading

Modules define functions and values. To execute code when a module is loaded,
put it inside a function with the name `.on_load`. This function is similar to
the hook for the `.onLoad` *package* namespace event.

This function is executed the first time the module is loaded in an R session.
Subsequent calls to `box::use` for that module, regardless of whether they occur
in a different scope, will refer to the already loaded, cached module, and will
*not* reload the module.

We can illustrate this by loading a module which has side-effects, `info`.

```{r file = 'info.r'}
```

Let’s use it:

```{r}
box::use(./info)
```

We have imported the module, and get the diagnostic messages. Let’s re-use the
module:

```{r}
box::use(./info)
```

… no messages are displayed. However, we *can* explicitly reload a module. This
clears the cache, and loads the module again. This can be useful during
development and debugging:

```{r}
box::reload(info)
```

And this displays the messages again. The `reload` function is a shortcut for
`unload` followed by `import` (using the exact same arguments as used on the
original `import` call).

### Module helper functions

This `info` module also show-cases two important helper functions:

1. `box::name` returns the name of the module with which it was loaded. This is
   especially handy because, when called outside of a module, `box::name` is
   `NULL`. This allows testing whether a piece of code was loaded as a module,
   or invoked directly (e.g. via `Rscript` on the command line).

2. `box::file` is similar to `system.file`: it returns the full path to any file
   within the directory where a module is stored. This is useful when
   distributing data files with modules, which are loaded from within the
   module. When invoked without arguments, `box::file` returns the full path to
   the directory containing the module source file.
