---
title: "Using compiled code"
author: Konrad Rudolph
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
    md_document:
        variant: gfm
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{Using compiled code}
    %\VignetteEncoding{UTF-8}
---
```{r include = FALSE}
devtools::load_all()
box::use(./source_file)

# Set up a temporary directory with the module demo. This is necessary for the
# sole reason that, otherwise, `R CMD check` breaks: vignette building invokes
# `tools::buildVignette`, which cleans up after itself by checking for new files
# and removing those. But creating a file inside a folder changes the folder
# timestamp, and hence the whole folder is removed.
tmpdir = tempfile('dir')
dir.create(tmpdir)
file.copy(file.path(getwd(), 'c'), tmpdir, recursive = TRUE)
box::set_script_path(file.path(tmpdir, 'c'))
```

## Current status

The ‘box’ package doesn’t have a built-in foreign function interface yet but it
is possible to integrate compiled code via R’s `SHLIB` mechanism for building
shared libraries. In particular, this also works with packages such as [Rcpp][].

For the time being, the following steps unfortunately require manual work:

1. Compile code when a module is installed or loaded for the first time.
2. Load the compiled shared library when a module is loaded.
3. Interface with compiled code via function calls.

This guide aims to describe all steps in sufficient detail to make them easy. In
the long run, the plan is to automate all these steps.

## Example

To demonstrate these steps, we will use an example module named `c` that uses
compiled code written in C. Here is the C code that we want to make usable in an
R module, which is saved in the file `c/hello.c`:

```{r file = 'c/hello.c'}
```

In addition, compiled code often includes specific compilation instructions.
While this is unnecessary for this simple example, it’s included anyway for
completeness. For R, these compilation instructions are contained in a file
called [`Makevars`][Makevars]:

```{r file = 'c/Makevars'}
```

## Compile code

To make code loadable and callable by R, it should be compiled via the `R CMD
SHLIB` mechanism. This only needs to happen *once* for each module, when it is
first loaded. By convention, such code should go into a submodule called
`__setup__`. This convention makes it clear that this is a “special” module,
and not intended for direct consumption of the module user.

The module specifies which object files to compile, and invokes `R CMD SHLIB`:

```{r file = 'c/__setup__.r'}
```

In principle, only the last line in this file should need to be changed for
other C projects.

With this in place, we can invoke the compilation by loading the `c/__setup__`
submodule:

```{r}
box::use(./c/`__setup__`)
```

The result of the compilation will be a single file, `hello.so` (on Unix and
macOS) or `hello.dll` (on Windows) which represents a *shared library file*, and
which we can load and use inside R.

## Loading compiled code

Compiled code from a shared library is loaded in R using the `dyn.load` command.
This will happen inside the module that uses and/or exposes the compiled code.

Since the name of the shared library file is platform dependent, we need a
helper function that gives us this name:

```{r eval = FALSE}
libname = function (name) {
    box::file(paste0(name, .Platform$dynlib.ext))
}
```

Now our module can load the compiled code; since this code needs to be executed
every time the module is loaded, it goes into the `.on_load` hook:

```{r eval = FALSE}
.on_load = function (ns) {
    ns$dll = dyn.load(libname('hello'))
}
```

… and don’t forget to unload the dynamic library when the module is unloaded:

```{r eval = FALSE}
.on_unload = function (ns) {
    dyn.unload(libname('hello'))
}
```


## Interacting with compiled code

Finally, our module needs a way of calling the compiled code. This is done via
the R primitive `.Call`:

```{r}
#' @export
hello_world = function (name) {
    .Call(dll$hello_world, name)
}
```

To use the code, we load the module and call the `hello_world` function:

```{r}
box::use(./c)
c$hello_world('Rthur')
```

Note that using `dll$hello_world` causes a somewhat costly call to
`getNativeSymbolInfo` every time the function is invoked. If this is undesired,
the value of `dll$hello_world` should be stored in a variable when loading the
module.

[Rcpp]: https://cran.r-project.org/package=Rcpp
[Makevars]: https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Using-Makevars
