---
title: "Module access benchmark"
author: Konrad Rudolph
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
    md_document:
        variant: gfm
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{Module access benchmark}
    %\VignetteEncoding{UTF-8}
---

<div class="dev-note">

> **Note:** This document describes internal implementation details. They are not required for users of the ‘box’ package and module authors.

</div>


## Purpose

Accessing a name in a module via `$` is expected to be an *extremely common* operation, and will happen in tight loops. Performance is therefore crucial. Furthermore, <span class="pkg">box</span> overrides `$` for modules to prevent partial matching (which the base R implementation does) and to provide clean, relevant, informative error messages when a nonexistent name is accessed.

This benchmark compares several implementations to choose the most efficient one that still fulfils the design criteria.


## Scaffold

The following code generates mock modules for lookup with few and many exported names as two representative cases. In fact, I don’t expect there to be a big difference between them since even large modules will typically have compartively few names (it might get more interesting with thousands of names, but this isn’t a case we expect often, and therefore don’t optimise for it).

And in fact using a hash table representation for typical module sizes probably *adds* overhead; so we also test non-hashed environments.

We will arbitrarily use `utils::tail` as the value that is exported from the test modules. The reason for this is so that we can compare performance with that of a package export (via `::`) and still allow the <span class="pkg">bench</span> package to validate that all expressions return the same result (via implicit `check = TRUE`).

```{r}
create_module = function (class, size, hash) {
    extra_objects = switch(
        size,
        small = list(),
        # 50 arbitrary names ought to do:
        large = as.list(setNames(nm = c(month.name, month.abb, letters)))
    )
    objects = c(extra_objects, name = utils::tail)
    structure(
        list2env(objects, parent = emptyenv(), hash = hash),
        class = if (class != 'baseline') class
    )
}
```

Some setup for compiling and loading native code:

```{r}
load_native = function (name) {
    oldwd = setwd('extract')
    on.exit(setwd(oldwd))

    exitcode = system2('R', c('CMD', 'SHLIB', paste0(name, '.c')))
    stopifnot(exitcode == 0L)

    dll_path = file.path(getwd(), paste0(name, .Platform$dynlib.ext))
    getNativeSymbolInfo(name, dyn.load(dll_path))
}
```


## Implementations

### `[[` extract

This serves as a comparison, but the semantics are wrong: it does not raise an error for nonexistent names; instead it returns `NULL`:

```{r}
`$.mod_brackets` = function (e1, e2) {
    e1[[e2, exact = TRUE]]
}
```


### `get`

The simplest compliant implementation, though it creates terrible error messages:

```{r}
`$.mod_get` = function (e1, e2) {
    get(e2, envir = e1, inherits = FALSE)
}
```


### `get` + `tryCatch`

The following implementations avoid the error raised by `get` for a nonexistent name and provide a better error message. The messages below are placeholders — a real implementation would trade time and complexity for a nicer error message (the performance of the error case is irrelevant for us, so it is permitted to be slow).

The first implementation catches the error raised by `get` and raises a different error instead:

```{r}
`$.mod_try_catch_get` = function (e1, e2) {
    tryCatch(
        get(e2, envir = e1, inherits = FALSE),
        error = function (e) {
            stop('Nice error message')
        }
    )
}
```


### `get` + `%in% names(…)`

The next implementation adds an explicit check via lookup in the `names` of the module environment:

```{r}
`$.mod_in_names_get` = function (e1, e2) {
    if (! e2 %in% names(e1)) {
        stop('Nice error message')
    }
    get(e2, envir = e1, inherits = FALSE)
}
```


### `get` + `%in% ls(…)`

Instead of `names`, we can also use `ls` to enumerate an environment’s name, but we need to remember to pass `all.names = TRUE`:

```{r}
`$.mod_in_ls_get` = function (e1, e2) {
    if (! e2 %in% ls(e1, all.names = TRUE)) {
        stop('Nice error message')
    }
    get(e2, envir = e1, inherits = FALSE)
}
```


### `get` + `hasName`

`hasName(a, b)` does basically the same as `b %in% names(a)` but its documentation claims that it is more efficient:

```{r}
`$.mod_has_name_get` = function (e1, e2) {
    if (! hasName(e1, e2)) {
        stop('Nice error message')
    }
    get(e2, envir = e1, inherits = FALSE)
}
```


### `get` + `exists`

A more straightforward check for existence of a name in an environment is `exists`:

```{r}
`$.mod_exists_get` = function (e1, e2) {
    if (! exists(e2, envir = e1, inherits = FALSE)) {
        stop('Nice error message')
    }
    get(e2, envir = e1, inherits = FALSE)
}
```

### `get0`

It would be great if the `ifnotfound` argument to `get0` were lazily evaluated: we could pass the error handling logic directly to `get0`, to be evaluated only in the case of a nonexistent name. Alas, R does not indulge us. We therefore need to use a sentinel value for `ifnotfound` which is guaranteed to never be found in an actual module. Luckily this is straightforward by using a new environment: R performs reference identity checking for environments when passed to `identical`, meaning that two environments compare identically if and only if they result from the same creation via `new.env` (in particular, two environments are *not* identical just because they happen to contain the same — or no — contents).

In the implementation below this environment is reused between multiple invocations. In theory, an enterprising user *could* extract this value from the innards of <span class="pkg">box</span> and export it from their own module. But this wouldn’t happen by accident, and they deserve what they get.

Luckily for us, checking whether two environments are identical is also efficient (R just compares their pointers):

```{r}
`$.mod_get0` = function (e1, e2) {
    ret = get0(e2, e1, inherits = FALSE, ifnotfound = mod_get0_sentinel)
    if (identical(ret, mod_get0_sentinel)) {
        stop('Nice error message')
    }
    ret
}

mod_get0_sentinel = new.env()
```


### `get0` with handler in `on.exit`

We can avoid a temporary variable for the return value of `get0` by moving the sentinel check into the `on.exit` handler. I find this rather elegant, but I also suspect that it would be a lot less efficient:

```{r}
`$.mod_on_exit_get0` = function (e1, e2) {
    on.exit(
        if (identical(returnValue(), mod_get0_sentinel)) {
            stop('Nice error message')
        }
    )
    get0(e2, e1, inherits = FALSE, ifnotfound = mod_get0_sentinel)
}
```


### `.Call` native code

Finally, we move to native code implementations. The actual lookup implementation is always the same:

```c
SEXP name = Rf_installTrChar(STRING_ELT(e2, 0));
SEXP ret = Rf_findVarInFrame(e1, name);
if (ret == R_UnboundValue) {
    // handle nonexistent name
}
return ret;
```

Where `e1` and `e2` refer to the module environment and the name, respectively.

The first implementation is an unadorned `.Call` dispatch:

```{r}
`$.mod_call` = function (e1, e2) {
    .Call(.c_call, e1, e2)
}

.c_call = load_native('call')
```


### `.Call` native code with `environment` argument

To generate nicer error messages, it is tempting to pass additional information to the call (in particular, the calling environment):

```{r}
`$.mod_call_env` = function (e1, e2) {
    .Call(.c_call_env, e1, e2, parent.frame())
}

.c_call_env = load_native('call_env')
```


### `.External` native code

An alternative way of invoking native code is via `.External`, which looks the same on the calling (R) side of things:

```{r}
`$.mod_external` = function (e1, e2) {
    .External(.c_external, e1, e2)
}

.c_external = load_native('external')
```

The only difference is that we need to manually unpack the arguments from the pairlist on the C side:

```c
SEXP e1 = CADR(args);
SEXP e2 = CADDR(args);
// etc.
```


### `.External` native code with `environment` argument

And once again, with an extra argument for better context in the error message:

```{r}
`$.mod_external_env` = function (e1, e2) {
    .External(.c_external_env, e1, e2, parent.frame())
}

.c_external_env = load_native('external_env')
```


## Benchmark

The following distinguishes “default”, where no class has been assigned to the module (and thus the built-in environment type is used) and where an S3 class is assigned but no `$` implement, so that the S3 dispatch falls back to the default implementation. All other methods are dispatched to a specific, custom method implementation of the `$` generic. For good measure we also compare to qualified name lookup in a package via `::`.

Before looking at the results, it is helpful to write down our expectations, based on intuition and knowledge of the implementation of the various methods. This will allow us to verify, to some extent, our mental model of the R evaluation:

“baseline” and “default” will probably be the fastest cases since all the work being done happens deep inside the R interpreter in native code. It remains to be seen if our own native code implementation can match their performance (*in principle* it could even surpass it, since it needs to do less work by not performing partial matching). There might also be a special case for “baseline” (where no S3 class is assigned), and R might not perform S3 dispatch in this case at all.

`[[` and `get` both directly invoke native code (`[[` via `.Primitive` and `get` via `.Internal`). However, `[[` can perform S3 dispatch so it might be somewhat slower. Then again, `get` has more arguments that might need to be evaluated eagerly.

Handling errors explicitly will inevitably add overhead. By looking at the implementations, `tryCatch` adds a fair bit of R code. On the opposite side, `names` is a primitive function, and `%in%` (via `match`) is an internal function; but, compared to `tryCatch`, the `%in%` check is performing a *redundant* test, so I have no clear expectation of what method should be faster. `ls`, on the other hand, contains a non-trivial amount of R logic, so I’d expect it to be strictly slower than `names`. The documentation of `hasName` claims that it is more efficient than `%in%` + `names` but given its implementation (in terms of `match` and `names`) I don’t see how that would be possible.

Of all the ways of redundantly checking for nonexistent names, `exists` should be the fastest since it is implemented as an `.Internal` function and is more direct than the other ways; in particular, it can take advantage of the internal hash table structure of the environment, which none of the other methods can; however, this should only be an advantage for unrealistically large modules. For real-world modules, using a hash table for the environment probably makes no difference for lookup (at worst it might actually *add* a small overhead compared to linear scanning!).

`get0` is interesting: the documentation states that it is more efficient than `exists` + `get`, but the documentation only addresses the case where `ifnotfound = NULL`, which cannot be used here (`NULL` is a legitimate value to export from a module!). Is it still more efficient when calling `identical` with an environment instead of `is.null`?

… using `on.exit` to handle the sentinel value is almost certainly going to be slower than doing it directly.

```{r}
labels = c(
    baseline = 'baseline',
    default = 'default',
    brackets = '[[',
    get = 'get',
    try_catch_get = 'get + tryCatch',
    in_names_get = 'get + %in% names',
    in_ls_get = 'get + %in% ls',
    has_name_get = 'get + hasName',
    exists_get = 'get + exists',
    get0 = 'get0',
    on_exit_get0 = 'get0 + on.exit',
    call = '.Call',
    call_env = '.Call + env',
    external = '.External',
    external_env = '.External + env'
)

classes = names(labels)

cases = expand.grid(
    size = c('small', 'large'),
    hash = c(TRUE, FALSE),
    stringsAsFactors = FALSE
)

hash_label = function (hash) {
    c('hash', 'vec')[match(hash, c(TRUE, FALSE))]
}

modules = unlist(lapply(
    seq_len(nrow(cases)),
    function (row) {
        size = cases$size[row]
        hash = cases$hash[row]

        setNames(
            lapply(paste0('mod_', classes), create_module, size, hash),
            paste0(classes, '_', size, '_', hash_label(hash))
        )
    }
))

invisible(list2env(modules, .GlobalEnv))

combined_labels = unlist(lapply(
    seq_len(nrow(cases)),
    function (row) {
        size = cases$size[row]
        hash = cases$hash[row]
        sprintf('%s;%s;%s', labels, size, hash_label(hash))
    }
))

exprs = setNames(
    lapply(names(modules), function (mod) bquote(.(as.name(mod))$name)),
    combined_labels
)
all_exprs = c(exprs, `package;;;` = quote(utils::tail))

bm = bench::mark(iterations = 1e5, memory = FALSE, time_unit = 'us', exprs = all_exprs)
```

```{r echo = FALSE}
bm[, c('method', 'size', 'hash')] = do.call(rbind, strsplit(names(bm$expression), ';'))
bm$`median [µs]` = bm$median

knitr::kable( bm[, c('method', 'size', 'hash', 'median [µs]', 'itr/sec')], digits = 2L)
```

Some observations:

```{r echo = FALSE}
env_rows = grep(' \\+ env', bm$method)
overheads = unclass(bm$median[env_rows] / bm$median[env_rows - 1L])
```

* R code is *slow*, so any method that executes lots of it is slower than the alternatives which execute less of it. This doesn’t come as a big surprise, but the extent is still striking.
* It is reassuring to see that `hasName` is in fact identical in performance to `%in%` + `names`, contrary to what its documentation claims.
* All methods are **substantially** slower than the built-in lookup. This is disappointing: it adds a non-negligible, unavoidable overhead to every single qualified name lookup.
* Worse, even the best implementation is an order of magnitude slower than package name lookup via `::`, despite the fact that the latter first needs to check whether the package is already loaded. It’s great that `::` is fast, it’s not great that we are prevented from getting the same performance.
* Adding more arguments to the native calls adds a significant overhead: it makes them around `r sprintf('%0.2f', mean(overheads))`±`r sprintf('%0.2f', sd(overheads))` (±SD) times slower, and this is true for both the `.Call` and the `.External` calling conventions, and it occurs even when the arguments are not used.
* In R code, using a hashed *vs.* unhashed environment makes virtually no difference, except for large modules when checking via `%in%` + `names`. However, in compiled code, using a hashed environment is slightly but significantly faster.


### Correctness check

Ensure all of them (except for the baseline cases) throw an error when an invalid name is queried:

```{r}
invalid_exprs = lapply(exprs, function (e) `[[<-`(e, 3L, as.name('nam')))
self = environment()
exclude = grepl('^(baseline|default|\\[\\[);', names(invalid_exprs))

did_call_stop = vapply(
    invalid_exprs[! exclude],
    function (e) tryCatch({eval(e, self); FALSE}, error = function (.) TRUE),
    logical(1L)
)

stopifnot(all(did_call_stop))
```


## Trailer

<details>

<summary><code>sessionInfo()</code> output</summary>

```{r}
sessionInfo()
```

</details>
