---
title: "Frequently asked questions"
author: Konrad Rudolph
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
    md_document:
        variant: gfm
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{Frequently asked questions}
    %\VignetteEncoding{UTF-8}
---

## General

### Can I call `library`/`require`/`source` inside a module?

No, this usually won’t work: `library` (and `require`) attaches names *globally*, rather than inside the calling module. Inside a module, `box::use` must be used instead.

Likewise, by default `source` evaluates the source code in the global rather than in a local environment. This behaviour can be changed by setting its `local` argument to a suitable value.

Calling these functions inside a module will raise a warning which can be silenced by setting `options(box.warn.legacy = FALSE)`. After changing this option, ‘box’ needs to be (unloaded and) reloaded for the change to take effect.

### Can I use “meta” packages like ‘tidyverse’ with ‘box’?

No. You can use *individual* packages from the tidyverse via `box::use`. But directly loading a “meta” package such as ‘tidverse’ is intentionally not supported.

“Meta” packages such as ‘tidyverse’ are antithetical to the purpose of ‘box’: the whole point of ‘box’ is to make package imports *explicit*. Using “meta” packages would undermine this and is therefore discouraged.


### How to organise globally installed modules?

Module names need to be *fully qualified*, meaning that using them requires providing a name consisting of a prefix (at least one high-level namespace) and the module’s proper name. The choice of prefix is, to some extent, arbitrary. However, there are some common conventions that are worth following.

For example, it is common practice to use a company or user name (look at code sharing websites such as GitHub for inspiration). A project consisting of several nested modules might also serve the same purpose. For example, if ‘dplyr’ were implemented as a module, its common fully qualified name would probably be `tidyverse/dplyr`.

As a concrete example, consider the following modules, stored in the module search path:

```
‹box.path›
├── klmr
│   ├── fun
│   ├── ggplots
│   └── sys
├── mschubert
│   └── ebits
└── …
```

Using the `sys` module (more precisely, `klmr/sys`) requires the following R code:

```{r eval = FALSE}
box::use(klmr/sys)
```


## Common error messages

### “object 'X' not found” inside a module

Inside modules, only the ‘base’ package is attached. *All* other packages need to be attached, including all core R packages that are otherwise attached by default. This includes ‘stats’, ‘graphics’, ‘grDevices’, ‘utils’, ‘datasets’ and ‘methods’.

To use names from any of these packages, the packages needs to be loaded via `box::use` same as third-party packages. Alternatively, the standard module `r/core` can be used (and, optionally, attached):

```{r eval = FALSE}
box::use(r/core[...])
```

This is equivalent to using and attaching the default R packages ‘methods’, ‘stats’, ‘graphics’, ‘grDevices’ and ‘utils’, and is therefore similar to the state of a regular R session.

This module ships with ‘box’ and is always findable, regardless of the value of the module search path.



### “there is no package called ‘X’” when running `box::use(X)`, even though the module 'X' exists

`box::use(name)` never attempts to load a module, it always attempts to load a *package* called ‘name’, and fails if no such package exists. Modules need to be either *local* and start with `.` or `..`, or they need to be fully qualified. This means that the full module name contains at least one separator (`/`).

In practice, this means that modules inside the global module search path (set via the environment variable `R_BOX_PATH` or via `getOption('box.path')`), modules need to be located in nested folders.

See [how to organise globally installed modules?](#how-to-organise-globally-installed-modules-) for details.
