---
title: "Testing modules"
author: Konrad Rudolph
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
    md_document:
        variant: gfm
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{Testing modules}
    %\VignetteEncoding{UTF-8}
---
```{r include = FALSE}
devtools::load_all()
box::use(./source_file)
```

## Testing is crucial

Writing tests to ensure code correctness is a crucial part of developing robust
software. This is especially true for dynamic languages such as R, which lack
some of the tools that ensure correctness in statically checked programming
languages.

‘box’ does not dictate a specific style of testing, and different kinds of
testing are appropriate in different situations. However, the following article
suggests a structure for the test accompanying a module which I have found to
work well.

Following the suggested structure causes the implementation and the tests of
each module to be located in paths next to each other. This structure is found
in several other conventions across programming languages; but it differs
somewhat from the convention of R packages (where all implementation code is in
a separate directory from all testing code), and differs drastically from the
convention found for instance in Java. That said, it should be possible to make
a Java-like test project work with ‘box’, too.


## Support for existing testing frameworks

‘box’ is agnostic regarding the testing framework. The following example employs
the widely used ‘[testthat][]’ unit testing package, but other frameworks exist,
and should also work.

This example uses the [`bio/seq`](bio/seq.r) module from the _[Getting
started][]_ vignette. To enable unit testing, this module contains the following
code at the very end:

```{r eval = FALSE}
if (is.null(box::name())) {
    box::use(./`__tests__`)
}
```

This allows us to run the tests by running the module source code from the
command line, e.g. via

```bash
Rscript bio/seq.r
```

… or inside an IDE such as RStudio by choosing the menu item “Tools” › “Jobs” ›
“Start Local Job…” ^[It may be tempting to instead use the “Source” button but
doing so is problematic; see the section “A note on RStudio and other IDEs”!].
This works because `box::name` returns the name of the module from which this
function is called. But if the function is invoked from code that wasn’t loaded
via `box::use` (as is the case here), its value is `NULL`. In other words,
`is.null(box::name())` is a way of testing whether the code currently being run
is loaded as a module, or executed directly.

The code inside the `if` imports the `__tests__` submodule: that’s where we put
the unit tests. The name `__tests__` is a convention. You’re free to choose a
different name, but I recommend sticking with this convention. Note that
`__tests__` is not a valid R variable name; that’s why it needs to be written in
backticks, i.e. the qualified local module name is ``./`__tests__` ``.

In this case, the `__tests__` submodule consists of a directory with the
following contents:

* `__tests__/`
    * [`__init__.r`](bio/__tests__/__init__.r)
    * [`helper-module.r`](bio/__tests__/helper-module.r)
    * individual test files

The `__init__.r` file corresponds closely to the file `tests/testthat.R` in a
standard R package structure. It loads ‘testthat’ and launches the tests:

```{r file = 'bio/__tests__/__init__.r'}
```

This first loads and attaches the ‘testthat’ package. Although attaching is not
strictly necessary, ‘testthat’ code is a lot more readable without cluttering
the code with explicit name qualifications.

Next it invokes `test_dir` and passes the tests’ directory via `box::file`
inside the module’s `.on_load` hook — remember that only declarations should be
at file level inside a module! All code execution should happen inside
functions.


The `helper-module.r` file is a ‘testthat’ helper; these are sourced
automatically by ‘testthat’ in the environment where the tests are run. We use
this mechanism to load our module in the test environment:

```{r file = 'bio/__tests__/helper-module.r'}
```

And, again, attaching isn’t strictly necessary here. Note also that, depending
on how the tests are run, this helper file might not be needed, since executing
the module script file itself already loads the module contents into the global
namespace; however, not all ways of loading the tests do this; for instance,
RStudio’s “Start Local Job” doesn’t. So having this helper is sometimes
necessary, and never hurts.

With this set-up, the actual unit test files look exactly like regular
‘testthat’ test files. For instance, here’s `__tests__/test-seq.r`:

```{r file = 'bio/__tests__/test-seq.r'}
```


### A note on RStudio and other IDEs

Most IDEs have an option to “Source” a local file. When doing this it may *seem*
as if the tests are correctly run, but this isn’t actually the case! This is
because `box::use` doesn’t reload code that has already been loaded previously;
instead, it uses an already loaded, cached version. This means that running the
tests via the “Source” button risks running an outdated version of the tests, or
the module, or both, after modifying their code.

To avoid this, always execute the test module in a new R session. In RStudio,
the easiest way of doing this is by running it as a job, via the menu “Tools” ›
“Jobs” › “Start Local Job…” (or using the option “Source as Local Job…” in the
“Source” drop-down).


## Test interfaces, not implementation details

One big difference between testing module code and testing package code is that,
with the testing structure laid out above, the testing code only sees the
module’s public interface, it does not get access to the internal module
implementation.^[Depending on how it’s invoked; but we shouldn’t make
assumptions. In particular, when invoked as a job in RStudio, the test module is
loaded via the test helper, and thus the module internals are not visible.]

This is *by design*: the idea is that we want to test the *observable behaviour*
rather than the (purely incidental) current implementation, which might be
changed. This is the way unit testing works in many other environments, and how
it is often recommended.

But I realise that this may not always be appropriate. Sometimes we *need* to
test implementation details. There are essentially two workarounds for this. For
the moment, I have not yet developed a strong preference for either of these
methods.

1. Put the tests into the module files themselves.

    ```{r eval = FALSE}
    # mymod.r

    this_works = function () TRUE

    if (is.null(box::name())) {
        box::use(testthat[...])

        # Define tests

        test_that('implementation detail X works', {
            expect_true(this_works())
        })

        # Invoke tests

        test_file(box::file())
    }
    ```

2. Get access to the private module namespace for testing. When loading a module
   with `box::use`, the module object has an attribute `namespace` that holds
   the private module namespace. It generally shouldn’t be accessed by client
   code, but access by the testing code for a module is entirely legitimate:

    ```{r eval = FALSE}
    # __tests__/test-something.r

    box::use(../mymod)

    impl = attr(mymod, 'namespace')

    test_that('implementation detail X works', {
        expect_true(impl$this_works())
    })
    ```


[testthat]: https://testthat.r-lib.org/
[Getting started]: box.html
