```{r include=FALSE}
options(import.path = 'inst/demo')

source_file = function (path, language) {
    if (missing(language) || is.null(language))
        language = regmatches(path, regexpr('[^.]*$', path))
    code = paste(readLines(path), collapse = '\n')
    sprintf('```%s\n%s\n```', language, code)
}

knit_hooks$set(file = function (before, options, envir)
    if (! before)
        source_file(file.path('inst/demo', options$file), options$lang)
)
```

# Using *modules* to generate reusable R components

## Foreign language interface

Modules don’t have a built-in foreign language interface yet but it is possible
to effortlessly integrate C++ code via the excellent [Rcpp][] package.

As an example, take a look at the `rcpp` module found under `inst/demo`; the
module consists of a C++ source file which is loaded inside the `__init__.r`
file:

```{r file='rcpp/__init__.r'}
```

Here’s the C++ code itself (the example is taken from the Rcpp documentation):

```{r file='rcpp/convolve.cpp'}
```

This module can be used like any normal module:

```{r}
rcpp = import('rcpp')
ls(rcpp)
rcpp$convolve(1 : 3, 1 : 5)
```

---

Unfortunately, this has a rather glaring flaw: the code is recompiled for each
new R session. In order to avoid this, we need to compile the code *once* and
save the resulting dynamic library. There’s no straightforward way of doing
this, but Rcpp wraps `R CMD SHLIB`.

For the time being, we manually need to trigger compilation by executing the
[`__install__.r`][install.r] file found in the `inst/demo/rcpp` module path.

Once that’s done, the actual module code is easy enough:

```{r file='rcpp/compiled.r'}
```

We can use it like any other module:

```{r}
compiled = import('rcpp/compiled')
compiled$convolve(1 : 3, 1 : 5)
```

[Rcpp]: http://cran.r-project.org/web/packages/Rcpp/index.html
[install.r]: rcpp/__install__.r
